Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-07-15 Manuscript/sec4-experiments.tex
更新了一些问题

\begin{enumerate}
    \item The matrix multiplication operator mm in GCN is mainly used for vertex calculation $\gamma$, 
    this operator is particularly time-consuming on the \textit{cph} dataset, 
    because the dimensionality of the vertex feature vector input by \textit{cph} is very high,
    making the matrix multiplication in the vertex calculation of Layer0 The calculation is very high.
    Mul is the multiplication operation in the edge calculation function $\phi$.
    Both scatter\_add and gather are used to implement the aggregation step $\Sigma$ in the edge calculation,
    where the former is used in the forward stage and the latter is used in backward Phase.
    The index\_select operator is used in the collect step of the edge calculation.
    For the GCN algorithm, the edge calculation related operators occupy the main time-consuming.
    The time-consuming among the operators is relatively average, and there is no particularly outstanding performance bottleneck.
    \item The most time consuming in GGNN is also the matrix multiplication mm operator,
    which is mainly used for the vertex calculation function $\gamma$. The scatter\_add, index\_select and gather operators
    are used for edge calculations. And the \_thnn\_fused\_gru\_cell is used in the backward calculation of GRU.
    GGNN because of vertices With the increase in calculation complexity, the time-consuming time of the mm operator has increased significantly.
    \item The four most time-consuming operators in GAT are all related to the edge calculation. 
    Mul, \_index\_put\_impl and sum are used to implement the edge calculation function $\phi$.
    The index\_select operator is used in the collect phase of the edge calculation. 
    The mm operator is used vertex calculation function $\gamma$.
    \item The most time-consuming matrix multiplication operator mm in GaAN is used for both edge calculation and edge calculation,
    where edge calculation is dominant. Mul and cat are used for the $\phi$ function in edge calculation.
\end{enumerate}

In terms of commonality, 
\textbf{The main time-consuming of GNN calculation is still matrix multiplication mm, multiplication of mul and other operators by the number of elements},
so it is very suitable for calculation with GPU. Although the calculation of the aggregate step in edge calculation is relatively simple,
but Because it involves data synchronization and non-regular calculations (the degree of different vertices is very different),
the related operators scatter\_add and gather still occupy a certain amount of time.
Although the collection step in the edge calculation does not have any calculations,
but the related calculations The index\_select operator still accounts for about 10\% of the time-consuming.
\textbf{aggregate step and collect step are one of the computational performance bottlenecks of all GNN training},
optimizing the corresponding operator will improve the training efficiency of all GNNs.

@@

\begin{itemize}
    \item \textbf{The performance bottleneck of GNN training is affected by the average degree of the dataset}. 
    Because the average degree of most real-world graphs is above 10 degrees \cite{network-repository}, 
    the performance bottleneck of GNN training will be concentrated on the edges Calculation section.
    \item \textbf{According to the calculation complexity of the edge calculation function $\phi$, the performance bottleneck of GNN in edge calculation is different}:
    \begin{itemize}
        \item If the computational complexity of $\phi$ is high, the performance bottleneck is concentrated on the basic operators used to implement $\phi$.
        Optimizing the implementation of the corresponding basic operators will improve the training performance of this type of GNN.
        Taking GAT as an example, The most time-consuming operator \_index\_put\_impl in GAT is mainly used in the backward stage of the softmax calculation ($\alpha^k_{ij}$) in $\phi$,
        which only involves data movement.
        The optimized softmax implementation on the GPU can Significantly reduce GAT training time.
        \item If the computational complexity of $\phi$ is low, the collect and aggregate steps in its edge calculations are bottlenecks in computational performance. The collect step only involves a large amount of data movement. The aggregate step calculation is relatively simple (such as sum/average/maximum) Value, etc.), but because it involves data synchronization and irregular calculation, its time-consuming is still significant.
         Optimizing the implementation of these two steps on the GPU will improve the training performance of this type of GNN.
    \end{itemize}
\end{itemize}

@@

It is worth noting that *the peak memory during GNN training far exceeds the memory usage of the dataset itself.
The ratio of the highest peak memory usage during our training compared to the memory usage after Data Load is
defined as the memory expansion ratio. Fig~\ref{fig:exp_memory_expansion_ratio} compares the memory expansion ratios
of different GNNs on different datasets. The expansion ratio of GCN is the lowest, between 5-14 times,
and the expansion ratio of GaAN is the highest, up to 101 times
\textbf{Very high expansion ratio severely limits the data scalability of GNN, making GPU unable to process large-scale graph datasets},
especially restricting GNN with high edge calculation complexity.

@@

Before the sampling technique, GNN training is full batch, that is, all the vertices and edges in the training set participate in the training and calculate the gradient at the same time.
Full batch training can ensure convergence, but the training overhead is large each time, resulting in slow convergence.
Inspired by the mini batch training method in stochastic gradient descent, a series of GNN sampling techniques are proposed.
The sampling technique decomposes the training of the full graph (ie epoch) into several batches, and
each batch uses only part of the graph vertices and edges to participate Training and performing gradient
updates greatly reduces the training time of each batch, allowing multiple rounds of gradient descent to
be performed within a fixed time, thereby accelerating convergence. The experiments in this section mainly
analyze the impact of sampling techniques on training performance.

In the current implementation of PyG, the GNN model parameters always reside on the GPU,
and the dataset resides in the main memory. When processing each epoch, the CPU samples
the graph dataset in the main memory and generates several batches,
Each batch is a small-scale subgraph of the dataset. When training each batch, 
PyG copies the subgraph data corresponding to the batch to the GPU's memory,
trains based on the subgraph and updates the model parameters according to the gradient.
Based on the sampling technology and SGD optimization technology, 
the evaluation of the model parameters is performed every several epochs. 
The evaluation can be carried out on the CPU or on the GPU. 
Therefore, the statistical data in the experiments in this section does not include the evaluation phase. 
Neighbor is selected in this section Sampler and Cluster Sampler are two typical graph sampling techniques for analysis.

Fig~\ref{fig:exp_sampling_minibatch_graph_info} shows how the size of the subgraph sampled in the two sampling techniques varies with the batch size.
For the Neighbor Sampler, the relative batch size is equal to the number of vertices sampled in the last layer of GNN relative to the top of the full graph The proportion of verticess.
For Cluster Sampler, the batch size is equal to the number of partitions sampled compared to the number of all partitions in the whole image.
\textbf{Neighbor Sampler is very sensitive to the increase in batch size*. As the batch size increases, 
the sampling subgraph The number of vertices/edges and the average degree both increase rapidly and tend to stabilize}.
\textbf{Cluster Sampler is less sensitive to the increase in batch size*, the number of vertices/average degree increases
linearly with the increase in batch size, and the number of edges is in the batch When the size is relatively small,
it also increases linearly.}

It is worth noting that \textbf{the average degree of the subgraph sampled is much lower than the average degree of the whole picture,
especially when the relative batch size is low}. Taking the Neighbor Sampler at a relative batch size of 6\% as an example,
The average degree of the \textit{amp} dataset is 31.1, but the average degree of the sampled subgraph is only 5.8,
which is much lower than the average degree of the full picture;
the average degree of the subgraph sampled by the Cluster Sampler is lower, only 3.0.
Fig~\ref{fig:exp_sampling_minibatch_degrees_distribution) shows the comparison of the distribution of the sampled subgraph with the original image.
The overall distribution of the samples sampled by the two sampling techniques is lower than that of the original image. 
The reason is that the distribution of vertices in the actual graph dataset follows a power-rate distribution, and a small number of vertices The degree is very high,
which improves the average degree. In the sampling process, because the sampling method will limit the neighborhood size of the vertex, thereby reducing the upper limit of the vertex degree,
so that the average degree drops significantly. Combined with the experimental results in section 4.2 The decrease in the average degree of sampling subgraph will increase the time-consuming
proportion of vertex calculation. For GGNN with high vertex calculation complexity, vertex calculation will replace edge calculation and become a performance bottleneck.

\begin{figure}
    \centering
    \subfloat[Neighbor Sampler]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_minibatch_realtive_graph_info_graphsage_gcn.png}} \\
    \subfloat[Cluster Sampler]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_minibatch_realtive_graph_info_cluster_gcn.png}}
    \caption{The size of the sampled subgraph changes with the batch size. Each batch size is sampled 50 times, and the error bar represents the standard deviation. The relative batch size is the ratio relative to the full image.)}
    \label{fig:exp_sampling_minibatch_graph_info}
\end{figure}

@@

Fig~\ref{fig:exp_sampling_batch_train_time} shows the change of the training time of each batch on the \textit{amc} and \textit{fli} datasets with the batch size after using the sampling technique.
For the Neighbor Sampler, only when the batch size is very small The time spent in the training phase is significantly reduced compared to full-batch training. 
When the batch size is particularly small, the training time is significantly reduced. However, because the sampling itself has additional overhead, 
and there is also an overhead in transferring the sampled subgraph to the GPU, the training time of the entire batch may exceed the full image. 
Training is time-consuming. For Cluster Sampler, the sampled subgraph is smaller than Neighbor Sampler under the same relative batch size,
which makes the effect of sampling technology more obvious. But when the relative batch size increases, 
the additional cost of sampling technology The increase is very obvious. When the relative batch size $\geq$ 25\%,
the total time of each batch of the sampling technique even exceeds the time of the full image training. 
Experiments show that \textbf{the implementation of sampling technology in PyG at this stage is very inefficient and the extra overhead is high}.
When the batch size is slightly larger, the additional cost of sampling technology accounts for more than 50\% of the total training cost of the entire batch.
\textbf{Sampling technology can reduce training time only on very small batches}.

\begin{figure}
    \centering
    \subfloat[Neighbor Sampler on amc]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_relative_batch_size_train_time_stack_graphsage_amazon-computers.png}}
    \subfloat[Neighbor Sampler on fli]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_relative_batch_size_train_time_stack_graphsage_flickr.png}} \\
    \subfloat[Cluster Sampler on amc]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_relative_batch_size_train_time_stack_cluster_amazon-computers.png}}
    \subfloat[Cluster Sampler on fli]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_relative_batch_size_train_time_stack_cluster_flickr.png}}
    \caption{The training time of each batch changes with the batch size. FULL means that the full image participates in the training phase (excluding evaluation).}
    \label{fig:exp_sampling_batch_train_time}
\end{figure}

However, \textbf{the advantage of the sampling technique is that it can greatly reduce the peak memory usage overhead,
making large-scale graph neural network training possible}. 
Fig~\ref{fig:exp_sampling_memory_usage} shows how the peak memory usage changes with the relative batch size during the training process after sampling technology is used.
After sampling sampling technology, peak memory usage is greatly reduced. 
The decrease of Cluster Sampler is greater than that of Neighbor Sampler.

>>>>

\begin{enumerate}
    \item The matrix multiplication operator mm in GCN is mainly used for vertex calculation $\gamma$, 
    this operator is particularly time-consuming on the \textit{cph} dataset, 
    because the dimensionality of the vertex feature vector input by \textit{cph} is very high,
    making the matrix multiplication in the vertex calculation of Layer0 The calculation is very high.
    Mul is the multiplication operation in the edge calculation function $\phi$.
    Both scatter\_add and gather are used to implement the aggregation step $\Sigma$ in the edge calculation,
    where the former is used in the forward stage and the latter is used in backward Phase.
    The index\_select operator is used in the collect step of the edge calculation.
    For the GCN algorithm, the edge calculation related operators occupy the main time-consuming.
    The time-consuming among the operators is relatively average, and there is no particularly outstanding performance bottleneck.
    \item The most time consuming in GGNN is also the matrix multiplication mm operator,
    which is mainly used for the vertex calculation function $\gamma$. The scatter\_add, index\_select and gather operators
    are used for edge calculations. And the \_thnn\_fused\_gru\_cell is used in the backward calculation of GRU.
    GGNN because of vertices With the increase in calculation complexity, the time-consuming time of the mm operator has increased significantly.
    \item The four most time-consuming operators in GAT are all related to the edge calculation. 
    Mul, \_index\_put\_impl and sum are used to implement the edge calculation function $\phi$.
    The index\_select operator is used in the collect phase of the edge calculation. 
    The mm operator is used vertex calculation function $\gamma$.
    \item The most time-consuming matrix multiplication operator mm in GaAN is used for both edge calculation and vertex calculation,
    where edge calculation is dominant. Mul and cat are used for the $\phi$ function in edge calculation.
\end{enumerate}

In terms of commonality, 
\textbf{The main time-consuming of GNN calculation is still matrix multiplication mm, elementwise multiplication mul},
so it is very suitable for calculation with GPU. Although the calculation of the aggregate step in edge calculation is relatively simple,
but it involves data synchronization and non-regular calculations (the degree of different vertices is very different),
the related operators scatter\_add and gather still occupy a certain amount of time.
Although the collection step in the edge calculation does not have any calculationss,
but the related calculations index\_select operator still accounts for about 10\% of the time-consuming.
\textbf{aggregate step and collect step are one of the computational performance bottlenecks of all GNN training},
optimizing the corresponding operator will improve the training efficiency of all GNNs.

@@

\begin{itemize}
    \item \textbf{The performance bottleneck of GNN training is affected by the average degree of the dataset}. 
    Because the average degree of most real-world graphs is above 10 degrees \cite{network-repository}, 
    the performance bottleneck of GNN training will be concentrated on the edges calculation section.
    \item \textbf{According to the calculation complexity of the edge calculation function $\phi$, the performance bottleneck of GNN in edge calculation is different}:
    \begin{itemize}
        \item If the computational complexity of $\phi$ is high, the performance bottleneck is concentrated on the basic operators used to implement $\phi$.
        Optimizing the implementation of the corresponding basic operators will improve the training performance of this type of GNN.
        Taking GAT as an example, The most time-consuming operator \_index\_put\_impl in GAT is mainly used in the backward stage of the softmax calculation ($\alpha^k_{ij}$) in $\phi$,
        which only involves data movement.
        The optimized softmax implementation on the GPU can significantly reduce GAT training time.
        \item If the computational complexity of $\phi$ is low, the collect and aggregate steps in its edge calculation are bottlenecks in computational performance. The collect step only involves a large amount of data movement. The aggregate step calculation is relatively simple (such as sum/average/maximum) Value, etc.), but because it involves data synchronization and irregular calculation, its time-consuming is still significant.
         Optimizing the implementation of these two steps on the GPU will improve the training performance of this type of GNN.
    \end{itemize}
\end{itemize}

@@

It is worth noting that \textbf{the peak memory during GNN training far exceeds the memory usage of the dataset itself}.
The ratio of the highest peak memory usage during our training compared to the memory usage after data Load is
defined as the memory expansion ratio. Fig~\ref{fig:exp_memory_expansion_ratio} compares the memory expansion ratios
of different GNNs on different datasets. The expansion ratio of GCN is the lowest, between 5-14 times,
and the expansion ratio of GaAN is the highest, up to 101 times.
\textbf{Very high expansion ratio severely limits the data scalability of GNN, making GPU unable to process large-scale graph datasets},
especially restricting GNN with high edge calculation complexity.

@@

Before the sampling technique, GNN training is full batch, that is, all the vertices and edges in the training set participate in the training and calculate the gradient at the same time.
Full batch training can ensure convergence, but the training overhead is large each time, resulting in slow convergence.
Inspired by the minibatch training method in stochastic gradient descent, a series of GNN sampling techniques are proposed.
The sampling technique decomposes the training of the full graph (i.e. epoch) into several batches, and
each batch uses only part of the graph vertices and edges to participate training and performing gradient
updates greatly reduces the training time of each batch, allowing multiple rounds of gradient descent to
be performed within a fixed time, thereby accelerating convergence. The experiments in this section mainly
analyze the impact of sampling techniques on training performance.

In the current implementation of PyG, the GNN model parameters always reside on the GPU,
and the dataset resides in the main memory. When processing each epoch, the CPU samples
the graph dataset in the main memory and generates several batches,
Each batch is a small-scale subgraph of the dataset. When training each batch, 
PyG copies the subgraph data corresponding to the batch to the GPU's memory,
trains based on the subgraph and updates the model parameters according to the gradient.
Based on the sampling technology and SGD optimization technology, 
the evaluation of the model parameters is performed every several epochs. 
The evaluation can be carried out on the CPU or on the GPU. 
Therefore, the statistical data in the experiments in this section does not include the evaluation phase. 
Neighbor Sampler and Cluster Sampler is selected in this section as they are two typical graph sampling techniques for analysis.

Fig~\ref{fig:exp_sampling_minibatch_graph_info} shows how the size of the subgraph sampled in the two sampling techniques varies with the batch size.
For Neighbor Sampler, the relative batch size is equal to the number of vertices sampled in the last layer of GNN relative to the top of the full graph the proportion of vertices.
For Cluster Sampler, the batch size is equal to the number of partitions sampled compared to the number of all partitions in the whole graph.
\textbf{Neighbor Sampler is very sensitive to the increase in batch size*. As the batch size increases, 
the sampling subgraph The number of vertices/edges and the average degree both increase rapidly and tend to stabilize}.
\textbf{Cluster Sampler is less sensitive to the increase in batch size*, the number of vertices/average degree increases
linearly with the increase in batch size, and the number of edges is in the batch When the size is relatively small,
it also increases linearly.}

It is worth noting that \textbf{the average degree of the subgraph sampled is much lower than the average degree of the whole graph,
especially when the relative batch size is low}. Taking Neighbor Sampler at a relative batch size of 6\% as an example,
The average degree of the \textit{amp} dataset is 31.1, but the average degree of the sampled subgraph is only 5.8,
which is much lower than the average degree of the full graph.
The average degree of the subgraph sampled by the Cluster Sampler is lower, only 3.0.
Fig~\ref{fig:exp_sampling_minibatch_degrees_distribution) shows the comparison of the distribution of the sampled subgraph with the original graph.
The vertex dregree distribution of subgraphs sampled by the two sampling techniques is lower than that of the original graph. 
The reason is that the degree distribution of vertex in the actual graph dataset follows a power-rate distribution and only a small number of vertices is very high,
which improves the average degree. In the sampling process, because the sampling method will limit the neighborhood size of the vertex, thereby reducing the upper limit of the vertex degree,
so that the average degree drops significantly. Combined with the experimental results in section 4.2, the decrease in the average degree of sampled subgraph will increase the time-consuming
proportion of vertex calculation. For GGNN with high vertex calculation complexity, vertex calculation will replace edge calculation and become a performance bottleneck.

\begin{figure}
    \centering
    \subfloat[Neighbor Sampler]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_minibatch_realtive_graph_info_graphsage_gcn.png}} \\
    \subfloat[Cluster Sampler]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_minibatch_realtive_graph_info_cluster_gcn.png}}
    \caption{The size of the sampled subgraph changes with the batch size. Each batch size is sampled 50 times, and the error bar represents the standard deviation. The relative batch size is the ratio relative to the full graph.)}
    \label{fig:exp_sampling_minibatch_graph_info}
\end{figure}

@@

Fig~\ref{fig:exp_sampling_batch_train_time} shows the change of the training time of each batch on the \textit{amc} and \textit{fli} datasets with the batch size after using the sampling technique.
For the Neighbor Sampler, the time spent in the training phase is significantly reduced compared to full-batch training only when the batch size is very small. 
When the batch size is particularly small, the training time is significantly reduced. However, because the sampling itself has additional overhead, 
and there is also an overhead in transferring the sampled subgraph to the GPU, the training time of the entire batch may exceed the full graph. 
training is time-consuming. For Cluster Sampler, the sampled subgraph is smaller than Neighbor Sampler under the same relative batch size,
which makes the effect of sampling technology more obvious. But when the relative batch size increases, 
the increment by additional cost of sampling technology is very obvious. When the relative batch size $\geq$ 25\%,
the total time of each batch of the sampling technique even exceeds the time of the full graph training. 
Experiments show that \textbf{the implementation of sampling technology in PyG at this stage is very inefficient and the extra overhead is high}.
When the batch size is slightly larger, the additional cost of sampling technology accounts for more than 50\% of the total training cost of the entire batch.
\textbf{Sampling technology can reduce training time only on very small batches}.

\begin{figure}
    \centering
    \subfloat[Neighbor Sampler on amc]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_relative_batch_size_train_time_stack_graphsage_amazon-computers.png}}
    \subfloat[Neighbor Sampler on fli]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_relative_batch_size_train_time_stack_graphsage_flickr.png}} \\
    \subfloat[Cluster Sampler on amc]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_relative_batch_size_train_time_stack_cluster_amazon-computers.png}}
    \subfloat[Cluster Sampler on fli]{\includegraphics[height=4cm]{figs/experiments/exp_sampling_relative_batch_size_train_time_stack_cluster_flickr.png}}
    \caption{The training time of each batch changes with the batch size. FULL means that the full graph participates in the training phase (excluding evaluation).}
    \label{fig:exp_sampling_batch_train_time}
\end{figure}

However, \textbf{the advantage of the sampling technique is that it can greatly reduce the peak memory usage overhead,
making large-scale graph neural network training possible}. 
Fig~\ref{fig:exp_sampling_memory_usage} shows how the peak memory usage changes with the relative batch size during the training process after sampling technology is used.
After sampling sampling technology, peak memory usage is greatly reduced and the decrease of Cluster Sampler is greater than that of Neighbor Sampler.
